---
title: "p8105_hw3_tp2806"
author: "Tejashree Prakash"
date: 2025-10-12
output: github_document
---

```{r}
#Keep the output clean with code and outputs 
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  echo = TRUE 
)
```


```{r setup, include=FALSE}
#Load libraries and data
library(tidyverse)
library(dplyr)
library(patchwork)
library(p8105.datasets)

data("instacart")
```

## Problem 1 


```{r}
#exploratory analysis 
total_row <- nrow(instacart)
summary <- dim(instacart)  

#examples of orders 
example_orders <- instacart %>%
  select(order_id, user_id, aisle, product_name, order_dow, order_hour_of_day) %>%
  sample_n(3)
```

The instacart data contains `r total_row` rows. The `r summary[1]` variables, or columns. Key variables include: order_id, which is the order ID for each order; product_name, defining the specific item that was ordered; and aisle, defining aisle number matching to a specific order

For example, user `r example_orders$user_id[1]` placed order `r example_orders$order_id[1]`
that included "`r example_orders$product_name[1]`" from the "`r example_orders$aisle[1]`" aisle
on day `r example_orders$order_dow[1]` at hour `r example_orders$order_hour_of_day[1]`.

```{r}
#distinct aisle 
aisle_num <- instacart %>%
  distinct(aisle_id, aisle) %>%
  nrow()

#highlight the most items ordered
aisle_counts <- instacart %>%
  group_by(aisle_id, aisle) %>%
  summarise(n_count = n(), .groups = "drop") %>% 
  slice_max(order_by = n_count, n = 5)
```

There are `r aisle_num` aisles. The top 5 aisles overall are `r paste(aisle_counts$aisle, collapse = ", ")`. 


## Making a plot that shows the number of items ordered in each aisle. 
```{r}
#make a plot
aisle_plot <- instacart %>%
  count(aisle) %>%
  filter(n > 10000) %>%
  arrange(desc(n)) %>%
  mutate(aisle = fct_reorder(aisle, n)) %>%
  ggplot(aes(x = aisle, y = n)) +
    geom_col(aes(fill = n)) +
  scale_fill_gradient(low = "lightblue", high = "blue") + 
  coord_flip() +
  labs(
    title = "Number of items ordered in each aisle (> 10000 items)", 
    x = "Aisle", 
    y = "Number of itemds ordered"
  ) + 
  theme_minimal()

aisle_plot
```


## Making a table that show popular items in three aisles. 
```{r}
#list frequency of three most tables in three aisles 
pop_items <- instacart %>%
  filter(aisle %in% c("baking ingredients", "dog food care", "packaged vegetables fruits")) %>%
  group_by(aisle, product_name) %>%
  summarise(order_count = n(), .groups = "drop") %>%
  group_by(aisle) %>%
  top_n(3, order_count) %>%
  arrange(aisle, desc(order_count))

knitr::kable(pop_items)
```

## Making a readable table that shows the mean hour of the day at which apples and ice cream are ordered on each day of the week. 
```{r}
apple_ice_cream_orders <- instacart %>%
  filter(product_name %in% c("Pink Lady Apples", "Coffee Ice Cream")) %>%
  group_by(product_name, order_dow) %>%
  summarise(mean_hour = mean(order_hour_of_day), .groups = "drop") %>%
  mutate(
    mean_time = format(
      as.POSIXct(mean_hour * 3600, origin = "1970-01-01", tz = "UTC"), 
      format = "%I:%M %p"
    ),
    order_dow = recode(order_dow, 
                            `0` = "Sunday", 
                            `1` = "Monday",
                            `2` = "Tuesday",
                            `3` = "Wednesday",
                            `4` = "Thursday", 
                            `5` = "Friday", 
                            `6` = "Saturday")
                  ) %>%
  select(product_name, order_dow, mean_time) %>% #to ensure the pivot occurs without mean_hour variable
  pivot_wider( #makes it readable 
    names_from = "order_dow",
    values_from = "mean_time"
  ) 
knitr::kable(apple_ice_cream_orders)
```


## Problem 2

Downloading and cleaning the Zillow data (ZORI and zip code data).

```{r}
#Zip metadata
zip_df <- 
  read_csv("zillow_data/Zip Codes.csv") %>%
  janitor::clean_names() %>%
  mutate(file_date = as.Date(file_date, format = "%m/%d/%y")) #consistent date format

#Zillow data
zori_df <- 
  read_csv("zillow_data/zori.csv") %>%
  janitor::clean_names() %>%
  rename(zip_code = region_name, county = county_name) %>%
  pivot_longer(
    cols = starts_with("x20"),
    names_to = "date",
    values_to = "rent_price"
  ) %>%
  mutate(
    county = str_remove(county, " County$"),
    date = str_remove(date, "^x"), #remove "x"
    date = ymd(date), #consistent date format
  ) %>%
  rename(borough = county) 

#Merging datasets
zillow_df <- left_join(zori_df, zip_df, by= c("zip_code"), relationship =
  "many-to-many")

zillow_df <- zillow_df %>%
  select(date, zip_code, neighborhood, borough, city, state, rent_price) %>%
  mutate(
    borough = case_match(
      borough,
      "Bronx" ~ "Bronx",
      "Kings" ~ "Brooklyn",
      "New York" ~ "Manhattan",
      "Queens" ~ "Queens",
      "Richmond" ~ "Staten Island"
    ))
knitr::kable(head(zillow_df, 5))

```

```{r}
#summarize frequency of zip codes
zip_obs <- zillow_df %>%
  group_by(zip_code) %>%
  summarise(month_count = n(), .groups = "drop")

#how many zip codes observed 116 times
zip_116 <- zip_obs %>%
  filter(month_count == 116) %>%
  nrow()

#how many zip codes observed fewer than 10 times
zip_10 <- zip_obs %>%
  filter(month_count < 10) %>%
  nrow()

```
There are `r zip_116` zip codes that are observed 116 times. There are `r zip_10` zip codes that are observed fewer than 10 times. This suggests that there are not many zip codes that contain a limited number of rentals. This could be due to several reasons such as if there is greater property ownership rather than renting in that zip code area or the zip code doesn't contain land to be developed or hasn't been developed yet into homes (rural areas). 



## Creating a table that shows the average rental price in each borough and year. 
```{r}
rent_pricing <- zillow_df %>%
  mutate(year = year(date)) %>%
  group_by(borough, year) %>%
  summarise(
    avg_rent_price = mean(rent_price, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from = year,
    values_from = avg_rent_price
  ) 

knitr::kable(head(rent_pricing))
```
Across all boroughs/counties, there is an increase in average rent pricing in later years (2024) compared to earlier years (2015). Out of the boroughs, Manhattan has the largest increase in average rent price between 2015 to 2024. "NaN" refers to missing values where we don't have rent price data. 


## Making a plot that shows NYC rental prices within zip codes for all available years, comparing prices across boroughs. 

```{r}
rent_by_zip_df <- zillow_df %>%
  mutate(year = year(date)) %>% #get year from date
  group_by(borough, zip_code, year) %>%
  summarise(
    avg_rent_price = mean(rent_price, na.rm = TRUE), .groups = "drop"
  )
  
rent_by_zip_plot <- 
  ggplot(rent_by_zip_df, aes(x = year, y = avg_rent_price, group = zip_code, color = borough)) +
  geom_point(alpha = 0.75) +
  labs(
    title = "NYC Rental Prices by ZIP Code and Borough Over Time",
    x = "Year", 
    y = "Average Rent Price (USD)", 
    color = "Borough"
  ) +
  scale_x_continuous(
    breaks = c(2015, 2016, 2017, 2018, 2019, 2020, 2021, 2022, 2023, 2024),  
    labels = c(2015, 2016, 2017, 2018, 2019, 2020, 2021, 2022, 2023, 2024)
  )
print(rent_by_zip_plot)
```
This plot shows the average rent prices of each zip code within all boroughs and all years between 2015 to 2024. Overall, we can see through this plot that rent prices consistently ranged between 2000-4000 USD between 2015 to 2021. The cost of Manhattan rent is consistently higher than other boroughs throughout all years. Starting in 2022, we see a shift to more rental prices above 4000 USD, especially in Manhattan. This reflects inflation and other living costs raising after the COVID-19 pandemic. 



## Making visualization that shows average rental price over each month in 2023. 

```{r}
#Create table that shows average rental price over 2023
rent_2023 <- zillow_df %>%
  mutate(
    year = year(date), #converting year properly 
    month = month(date, label = TRUE, abbr = FALSE)
    ) %>%
  filter(year == 2023) %>% #filter to only 2023
  group_by(borough, zip_code, month) %>%
  summarise(
    avg_rent_price = mean(rent_price, na.rm = TRUE),
    .groups = "drop"
  ) 

#Create plot showing the distribution 
rent_2023_plot <- ggplot(rent_2023, aes(x = borough, y = avg_rent_price, fill = borough)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  facet_wrap(~ month, ncol = 3) +
  labs(
    title = "Distribution of Average Rent Prices by ZIP-Code and Borough, in 2023",
    x = "Borough",
    y = "Average Rent Price (USD)"
  ) +
  theme(axis.text.x = element_blank()) 

print(rent_2023_plot)
```
These box plots show the distribution of average rent prices within each zip code over each month in 2023, grouped by borough. Consistently across each month in 2023, Manhattan had the highest average rent price. Visually, Brooklyn seemed consistent in the average rent price across all of the months. Staten Island had the smallest distribution, due to the smaller amount of zip codes and land area compared to the other boroughs in NYC. 


## Both plots put together. 
```{r, fig.width=15, fig.height=12}
#Merge both plots into one graphic
rent_graphic <- rent_by_zip_plot + rent_2023_plot
print(rent_graphic)

#Save graphic into results folder
ggsave("results/rent_graphic.png", rent_graphic, width = 15, height = 10, dpi = 300)
```

## Problem 3

```{r}

```

