---
title: "p8105_hw3_tp2806"
author: "Tejashree Prakash"
date: 2025-10-12
output: github_document
---

```{r}
#Keep the output clean with code and outputs 
knitr::opts_chunk$set(
  message = FALSE,
  echo = TRUE 
)
```


```{r setup, include=FALSE}
#Load libraries and data
library(tidyverse)
library(dplyr)
library(patchwork)
library(p8105.datasets)

data("instacart")
```

## Problem 1 


```{r}
#exploratory analysis 
total_row <- nrow(instacart)
summary <- dim(instacart)  

#examples of orders 
example_orders <- instacart %>%
  select(order_id, user_id, aisle, product_name, order_dow, order_hour_of_day) %>%
  sample_n(3)
```

The instacart data contains `r total_row` rows. The `r summary[1]` variables, or columns. Key variables include: order_id, which is the order ID for each order; product_name, defining the specific item that was ordered; and aisle, defining aisle number matching to a specific order

For example, user `r example_orders$user_id[1]` placed order `r example_orders$order_id[1]`
that included "`r example_orders$product_name[1]`" from the "`r example_orders$aisle[1]`" aisle
on day `r example_orders$order_dow[1]` at hour `r example_orders$order_hour_of_day[1]`.

```{r}
#distinct aisle 
aisle_num <- instacart %>%
  distinct(aisle_id, aisle) %>%
  nrow()

#highlight the most items ordered
aisle_counts <- instacart %>%
  group_by(aisle_id, aisle) %>%
  summarise(n_count = n(), .groups = "drop") %>% 
  slice_max(order_by = n_count, n = 5)
```

There are `r aisle_num` aisles. The top 5 aisles overall are `r paste(aisle_counts$aisle, collapse = ", ")`. 


## Making a plot that shows the number of items ordered in each aisle. 
```{r}
#make a plot
aisle_plot <- instacart %>%
  count(aisle) %>%
  filter(n > 10000) %>%
  arrange(desc(n)) %>%
  mutate(aisle = fct_reorder(aisle, n)) %>%
  ggplot(aes(x = aisle, y = n)) +
    geom_col(aes(fill = n)) +
  scale_fill_gradient(low = "lightblue", high = "blue") + 
  coord_flip() +
  labs(
    title = "Number of items ordered in each aisle (> 10000 items)", 
    x = "Aisle", 
    y = "Number of itemds ordered"
  ) + 
  theme_minimal()

aisle_plot
```
The most ordered items in each aisle were fresh vegetables and fruits, packaged vegetables and fruits, yogurt, cheese, water, and milk. These orders make sense as that is the baseline of many people's shopping list. 

## Making a table that show popular items in three aisles. 
```{r}
#list frequency of three most tables in three aisles 
pop_items <- instacart %>%
  filter(aisle %in% c("baking ingredients", "dog food care", "packaged vegetables fruits")) %>%
  group_by(aisle, product_name) %>%
  summarise(order_count = n(), .groups = "drop") %>%
  group_by(aisle) %>%
  top_n(3, order_count) %>%
  arrange(aisle, desc(order_count))

knitr::kable(pop_items)
```
For baking ingredients, the most popular items were light brown sugar, pure baking soda, and cane sugar. For dog food/care, it was two different brands of dog treats/food and small dog biscuits. For packaged vegetables and fruits, organic baby spinach, raspberries, and blueberries were the most bought items. 


## Making a readable table that shows the mean hour of the day at which apples and ice cream are ordered on each day of the week. 
```{r}
apple_ice_cream_orders <- instacart %>%
  filter(product_name %in% c("Pink Lady Apples", "Coffee Ice Cream")) %>%
  group_by(product_name, order_dow) %>%
  summarise(mean_hour = mean(order_hour_of_day), .groups = "drop") %>%
  mutate(
    mean_time = format(
      as.POSIXct(mean_hour * 3600, origin = "1970-01-01", tz = "UTC"), 
      format = "%I:%M %p"
    ),
    order_dow = recode(order_dow, 
                            `0` = "Sunday", 
                            `1` = "Monday",
                            `2` = "Tuesday",
                            `3` = "Wednesday",
                            `4` = "Thursday", 
                            `5` = "Friday", 
                            `6` = "Saturday")
                  ) %>%
  select(product_name, order_dow, mean_time) %>% #to ensure the pivot occurs without mean_hour variable
  pivot_wider( #makes it readable 
    names_from = "order_dow",
    values_from = "mean_time"
  ) 
knitr::kable(apple_ice_cream_orders)
```
The average hour-of-day that coffee ice cream and pink lady apples were bought varied each day, but they were mostly contained in the afternoon hours. 

## Problem 2

Downloading and cleaning the Zillow data (ZORI and zip code data), from HW2.

```{r}
#Zip metadata
zip_df <- 
  read_csv("data/zillow_data/Zip Codes.csv") %>%
  janitor::clean_names() %>%
  mutate(file_date = as.Date(file_date, format = "%m/%d/%y")) #consistent date format

#Zillow data
zori_df <- 
  read_csv("data/zillow_data/zori.csv") %>%
  janitor::clean_names() %>%
  rename(zip_code = region_name, county = county_name) %>%
  pivot_longer(
    cols = starts_with("x20"),
    names_to = "date",
    values_to = "rent_price"
  ) %>%
  mutate(
    county = str_remove(county, " County$"),
    date = str_remove(date, "^x"), #remove "x"
    date = ymd(date), #consistent date format
  ) %>%
  rename(borough = county) 

#Merging datasets
zillow_df <- left_join(zori_df, zip_df, by= c("zip_code"), relationship =
  "many-to-many")

zillow_df <- zillow_df %>%
  select(date, zip_code, neighborhood, borough, city, state, rent_price) %>%
  mutate(
    borough = case_match(
      borough,
      "Bronx" ~ "Bronx",
      "Kings" ~ "Brooklyn",
      "New York" ~ "Manhattan",
      "Queens" ~ "Queens",
      "Richmond" ~ "Staten Island"
    ))
knitr::kable(head(zillow_df, 5))

```

```{r}
#summarize frequency of zip codes
zip_obs <- zillow_df %>%
  group_by(zip_code) %>%
  summarise(month_count = n(), .groups = "drop")

#how many zip codes observed 116 times
zip_116 <- zip_obs %>%
  filter(month_count == 116) %>%
  nrow()

#how many zip codes observed fewer than 10 times
zip_10 <- zip_obs %>%
  filter(month_count < 10) %>%
  nrow()
```
There are `r zip_116` zip codes that are observed 116 times. There are `r zip_10` zip codes that are observed fewer than 10 times. This suggests that there are not many zip codes that contain a limited number of rentals. This could be due to several reasons such as if there is greater property ownership rather than renting in that zip code area or the zip code doesn't contain land to be developed or hasn't been developed yet into homes (rural areas). Rental prices can also fluctuate throughout the year because of seasonal renting. For example, if one area is very cold in the winter months, it likely will have less renters as some people have other homes during the winter months. 



## Creating a table that shows the average rental price in each borough and year. 
```{r}
rent_pricing <- zillow_df %>%
  mutate(year = year(date)) %>%
  group_by(borough, year) %>%
  summarise(
    avg_rent_price = mean(rent_price, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from = year,
    values_from = avg_rent_price
  ) 

knitr::kable(head(rent_pricing))
```
Across all boroughs/counties, there is an increase in average rent pricing in later years (2024) compared to earlier years (2015). Out of the boroughs, Manhattan has the largest increase in average rent price between 2015 to 2024. "NaN" refers to missing values where we don't have rent price data. 


## Making a plot that shows NYC rental prices within zip codes for all available years, comparing prices across boroughs. 

```{r}
rent_by_zip_df <- zillow_df %>%
  mutate(year = year(date)) %>% #get year from date
  group_by(borough, zip_code, year) %>%
  summarise(
    avg_rent_price = mean(rent_price, na.rm = TRUE), .groups = "drop"
  )
  
rent_by_zip_plot <- 
  ggplot(rent_by_zip_df, aes(x = year, y = avg_rent_price, group = zip_code, color = borough)) +
  geom_point() +
  labs(
    title = "NYC Rental Prices by ZIP Code and Borough Over Time",
    x = "Year", 
    y = "Average Rent Price (USD)", 
    color = "Borough"
  ) +
  scale_x_continuous(
    breaks = c(2015, 2016, 2017, 2018, 2019, 2020, 2021, 2022, 2023, 2024),  
    labels = c(2015, 2016, 2017, 2018, 2019, 2020, 2021, 2022, 2023, 2024)
  )
print(rent_by_zip_plot)
```
This plot shows the average rent prices of each zip code within all boroughs and all years between 2015 to 2024. Overall, we can see through this plot that rent prices consistently ranged between 2000-4000 USD between 2015 to 2021. The cost of Manhattan rent is consistently higher than other boroughs throughout all years. Starting in 2022, we see a shift to more rental prices above 4000 USD, especially in Manhattan. This reflects inflation and other living costs raising after the COVID-19 pandemic. 



## Making visualization that shows average rental price over each month in 2023. 

```{r, fig.width=10, fig.height=12}
#Create table that shows average rental price over 2023
rent_2023 <- zillow_df %>%
  mutate(
    year = year(date), #converting year properly 
    month = month(date, label = TRUE, abbr = FALSE)
    ) %>%
  filter(year == 2023) %>% #filter to only 2023
  group_by(borough, zip_code, month) %>%
  summarise(
    avg_rent_price = mean(rent_price, na.rm = TRUE),
    .groups = "drop"
  ) 
knitr::kable(head(rent_2023))

#Create plot showing the distribution 
rent_2023_plot <- ggplot(rent_2023, aes(x = borough, y = avg_rent_price, fill = borough)) +
  geom_boxplot() +
  facet_wrap(~ month, ncol = 3) +
  labs(
    title = "Distribution of Average Rent Prices by ZIP-Code and Borough, in 2023",
    x = "Borough",
    y = "Average Rent Price (USD)"
  ) +
  theme(axis.text.x = element_blank()) 

print(rent_2023_plot)
```

These box plots show the distribution of average rent prices within each zip code over each month in 2023, grouped by borough. Consistently across each month in 2023, Manhattan had the highest average rent price. Visually, Brooklyn seemed consistent in the average rent price across all of the months. Staten Island had the smallest distribution, due to the smaller amount of zip codes and land area compared to the other boroughs in NYC. 


## Both plots put together. 
```{r, fig.width=15, fig.height=12}
#Merge both plots into one graphic
rent_graphic <- rent_by_zip_plot + rent_2023_plot + 
  plot_layout(ncol = 2) + 
  plot_annotation(
    title = "NYC Rental Prices by Zip Code and Borough",
    theme = theme(plot.title = element_text(hjust = 0.25))
  )
print(rent_graphic)

#Save graphic into results folder
ggsave("results/rent_graphic.png", rent_graphic, width = 15, height = 12, dpi = 300)
```

## Problem 3

```{r}
#Load the two datasets
accel_df <- 
  read_csv("data/problem3_data/nhanes_accel.csv") %>%
  janitor::clean_names() 

covar_df <- 
  read_csv("data/problem3_data/nhanes_covar.csv", skip = 4) %>% #skip the first few rows that's data dictionary info 
  janitor::clean_names()

#Merge datasets - joining by seqn ID
nhanes_df <-
  left_join(covar_df, accel_df, by = join_by(seqn))

#Remove less than 21, if missing demographic data, recoding to not be numeric when appropriate
nhanes_df <- nhanes_df %>%
  drop_na() %>%
  filter(age >= 21) %>%
  mutate(
    sex = factor(sex, 
                 levels = c(1, 2),
                 labels = c("Male", "Female")), 
    education = factor(education, 
                       levels = c(1, 2, 3),
                       labels = c("Less than high school", "High school equivalent", "More than high school")))

knitr::kable(head(nhanes_df, 5))
```

## Produce a table and visualization for education category demographics. 
```{r}
#Table of number of men and women by education 
sex_education_df <- nhanes_df %>%
  count(education, sex) %>%
  pivot_wider(
    names_from = sex, 
    values_from = n, 
    values_fill = 0
  ) %>%
  rename(Education = education)

knitr::kable(head(sex_education_df, 5))

#Plot showing age by sex and education
age_sex_education_plot <- ggplot(nhanes_df, aes(x = age, fill = sex)) +
  geom_density(alpha = 0.4) +
  facet_wrap(~education) + 
  labs(
    title = "Age Distribution by Sex and Education",
    x = "Age (years)",
    y = "Density",       
    fill = "Sex"
    ) + 
  theme_minimal()
age_sex_education_plot

```
Through the education and sex table, there isn't significant variation between the number of women and men and the education level they have. The largest disparity is at the "high school equivalent", where there are more men than women that have this as their highest level of education. 

The density plot demonstrates how age and sex interact within level of education received. The density peaks for both sexes in the 60+ age range for those with less than high school. For those who have more than high school education, the opposite is occurring where there is greater density among younger age groups 20-40. Overall, the female sex group tends to have more prominent peaks compared to the male sex group. 


# Plot total activity by age, education, and sex.  
```{r}
#Create total activity variable 
nhanes_total <- nhanes_df %>%
  mutate(
    total_activity = rowSums(across(starts_with("min")), na.rm = TRUE) #sum the total minutes with every column starting with min
  ) %>%
  select(seqn, sex, age, bmi, education, total_activity, everything())
knitr::kable(head(nhanes_total, 5))

#Plot total activity against age, with education and sex, trend line 
total_activity_plot <- nhanes_total %>%
  ggplot(aes(x = age, y = total_activity, color = sex)) + 
  geom_point(alpha = 0.4) +
  geom_smooth(method = "loess", se = FALSE) + #includes trend line
  facet_wrap(~ education) + #puts education as level 
  labs(
    title = "Total Daily Activity versus Age, with Sex and Education",
    x = "Age (years)",
    y = "Total Activity (minutes total)"
  ) + theme_minimal()
total_activity_plot
```
The total activity variable sums all of the activities noted across different data collection points into one total number. When plotted, we see that as age increases, total daily activity decreases. Additionally, the female sex group tends to have higher minutes of daily activity in the two greater education groups. In the "less than high school", there is a peak in activity at around 60 years of age, and in "high school equivalent" groups, there is a peak in activity at age 40.


# Produce 3-panel plot with 24-hour activity. 
```{r}
#Summarize minutes into 24 hour activity 
total_activity_summary <- nhanes_total %>%
  pivot_longer(
    cols = starts_with("min"),
    names_to = "minute",
    names_prefix = "min",
    values_to = "activity_count"
  ) %>%
  mutate(
    minute = as.numeric(minute),
    hour = minute / 60
  ) %>%
  group_by(hour, sex, education) %>%
  summarise(
    mean_activity = mean(activity_count, na.rm = TRUE), #calculate mean activity by hour
    .groups = "drop"
  )

#Produce three panel plot 
activity_24hour_plot <- ggplot(total_activity_summary, aes(x = hour, y = mean_activity, color = sex)) +
  geom_line(size = 0.35) +
  facet_wrap(~ education, ncol = 1) +
  labs(
    title = "Mean Physical Activity by Hour and Education Level",
    x = "Hour of Day",
    y = "Mean Activity Count",
    color = "Sex"
  ) +
  theme_minimal() +  
  scale_x_continuous(
    breaks = c(0, 4, 8, 12, 16, 20, 24),  
    labels = c("12AM", "4AM", "8AM", "12PM", "4PM", "8PM", "12AM")
  )
activity_24hour_plot
```
We summarized the activity data by hour to show how mean activity fluctuated throughout different points of the day, separated by sex and education level. The line plot demonstrates that physical activity peaked during the middle of the day, and stayed consistently high for a few hours until starting to decrease around 7pm. Physical activity is lowest between 12AM-7AM, which follows typical rhythms as people sleep and begin their days. 

Overall, those with less than high school education have slightly higher rates of physical activity. This could be due to this group having more physically demanding occupations where they move at a greater level compared to those with higher levels of education. There seems to be a slightly greater gap between male and female activity levels in the more than high school education group. 

