---
title: "p8105_hw3_tp2806"
author: "Tejashree Prakash"
date: 2025-10-12
output: github_document
---

```{r setup, include=FALSE}
#Load libraries and data
library(tidyverse)
library(dplyr)
library(patchwork)
library(p8105.datasets)

data("instacart")
```

## Problem 1 


```{r}
#exploratory analysis 
total_row <- nrow(instacart)
summary <- dim(instacart)  

#examples of orders 
example_orders <- instacart %>%
  select(order_id, user_id, aisle, product_name, order_dow, order_hour_of_day) %>%
  sample_n(3)
```

The instacart data contains `r total_row` rows. The `r summary[1]` variables, or columns. Key variables include: order_id, which is the order ID for each order; product_name, defining the specific item that was ordered; and aisle, defining aisle number matching to a specific order

For example, user `r example_orders$user_id[1]` placed order `r example_orders$order_id[1]`
that included "`r example_orders$product_name[1]`" from the "`r example_orders$aisle[1]`" aisle
on day `r example_orders$order_dow[1]` at hour `r example_orders$order_hour_of_day[1]`.

```{r}
#distinct aisle 
aisle_num <- instacart %>%
  distinct(aisle_id, aisle) %>%
  nrow()

#highlight the most items ordered
aisle_counts <- instacart %>%
  group_by(aisle_id, aisle) %>%
  summarise(n_count = n(), .groups = "drop") %>% 
  slice_max(order_by = n_count, n = 5)
```

There are `r aisle_num` aisles. The top 5 aisles overall are `r paste(aisle_counts$aisle, collapse = ", ")`. 


Making a plot that shows the number of items ordered in each aisle. 
```{r}
#make a plot
instacart_count <- instacart %>%
  group_by(aisle_id, aisle) %>%
  summarise(aisle_count = n(), .groups = "drop") %>%
  filter(aisle_count > 10000) #only aisle orders with aisle count above 10000 items ordered 

instacart_plot <- ggplot(instacart_count, aes(x = aisle_id, y = aisle_count)) + 
  geom_point(aes(color = aisle), alpha = 0.5) + 
  labs(
    title = "Number of items ordered in each aisle", 
    x = "Aisle ID", 
    y = "Number of items ordered", 
    color = "Item type"
  ) + 
  scale_x_continuous(
    breaks = c(0, 20, 40, 60, 80, 100, 120, 140),  
    labels = c(0, 20, 40, 60, 80, 100, 120, 140)
  )
instacart_plot
```



Making a table that show popular items in three aisles. 
```{r}
#list frequency of three most tables in three aisles 
pop_items <- instacart %>%
  filter(aisle %in% c("baking ingredients", "dog food care", "packaged vegetables fruits")) %>%
  group_by(aisle, product_name) %>%
  summarise(order_count = n(), .groups = "drop") %>%
  group_by(aisle) %>%
  top_n(3, order_count) %>%
  arrange(aisle, desc(order_count))

knitr::kable(pop_items)
```

Making a readable table that shows the mean hour of the day at which apples and ice cream are ordered on each day of the week. 
```{r}
apple_ice_cream_orders <- instacart %>%
  filter(product_name %in% c("Pink Lady Apples", "Coffee Ice Cream")) %>%
  group_by(product_name, order_dow) %>%
  summarise(mean_hour = mean(order_hour_of_day), .groups = "drop") %>%
  mutate(order_dow = recode(order_dow, 
                            `0` = "Sunday", 
                            `1` = "Monday",
                            `2` = "Tuesday",
                            `3` = "Wednesday",
                            `4` = "Thursday", 
                            `5` = "Friday", 
                            `6` = "Saturday")) %>%
  pivot_wider( #makes it readable 
    names_from = "order_dow",
    values_from = "mean_hour"
  ) 
knitr::kable(apple_ice_cream_orders)
```


## Problem 2

Downloading and cleaning the Zillow data (ZORI and zip code data).

```{r}
#Zip metadata
zip_df <- 
  read_csv("zillow_data/Zip Codes.csv") %>%
  janitor::clean_names() %>%
  mutate(file_date = as.Date(file_date, format = "%m/%d/%y")) #consistent date format

#Zillow data
zori_df <- 
  read_csv("zillow_data/zori.csv") %>%
  janitor::clean_names() %>%
  rename(zip_code = region_name, county = county_name) %>%
  pivot_longer(
    cols = starts_with("x20"),
    names_to = "date",
    values_to = "rent_price"
  ) %>%
  mutate(
    county = str_remove(county, " County$"),
    date = str_remove(date, "^x"), #remove "x"
    date = ymd(date), #consistent date format
  ) %>%
  rename(borough = county)

#Merging datasets
zillow_df <- left_join(zori_df, zip_df, by= c("zip_code"), relationship =
  "many-to-many")

zillow_df <- zillow_df %>%
  select(date, zip_code, neighborhood, borough, city, state, rent_price) 
```

```{r}
#summarize frequency of zip codes
zip_obs <- zillow_df %>%
  group_by(zip_code) %>%
  summarise(month_count = n(), .groups = "drop")

#how many zip codes observed 116 times
zip_116 <- zip_obs %>%
  filter(month_count == 116) %>%
  nrow()

#how many zip codes observed fewer than 10 times
zip_10 <- zip_obs %>%
  filter(month_count < 10) %>%
  nrow()

```


Creating a table that shows the average rental price in each borough and year. 
```{r}
rent_pricing <- zillow_df %>%
  mutate(year = year(date)) %>%
  group_by(borough, year) %>%
  summarise(
    avg_rent_price = mean(rent_price, na.rm = TRUE),
    .groups = "drop"
  )
```
Across all boroughs/counties, there is an increase in average rent pricing in later years compared to earlier years. New York borough, which relates to Manhattan, has the largest increase in average rent price between 2015 to 2024. 


Making a plot that shows NYC rental prices within zip codes for all available years, comparing prices across boroughs. 

```{r}
rent_comparison_plot <- 
  ggplot(rent_pricing, aes(x = year, y = avg_rent_price, color = borough)) +
  geom_point(alpha = 0.75) +
  labs(
    title = "Average Rental Price by Borough over Time",
    x = "Year", 
    y = "Average Rent Price (USD)", 
    color = "Borough"
  ) +
  scale_x_continuous(
    breaks = c(2015, 2016, 2017, 2018, 2019, 2020, 2021, 2022, 2023, 2024),  
    labels = c(2015, 2016, 2017, 2018, 2019, 2020, 2021, 2022, 2023, 2024)
  )
```


Making visualization that shows rental price over each month in 2023. 

```{r}
#Create table that shows average rental price over 2023
rent_2023 <- zillow_df %>%
  mutate(
    year = year(date),
    month = month(date, label = TRUE, abbr = FALSE)
    ) %>%
  filter(year == 2023) %>% #filter to only 2023
  group_by(borough, zip_code, month) %>%
  summarise(
    avg_rent_price = mean(rent_price, na.rm = TRUE),
    .groups = "drop"
  ) 

#Create plot showing the distribution 
rent_2023_plot <- ggplot(rent_2023, aes(x = borough, y = avg_rent_price, fill = borough)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  facet_wrap(~ month, ncol = 3) +
  labs(
    title = "Distribution of Average ZIP-Code Rent Prices by Borough, in 2023",
    x = "Borough",
    y = "Average Rent Price (USD)"
  ) +
  theme(axis.text.x = element_blank()) +
  theme_minimal() 
```


Both plots put together. 
```{r}
#Merge both plots into one graphic
rent_graphic <- rent_comparison_plot + rent_2023_plot

#Save graphic into results folder
ggsave("results/rent_graphic.png", rent_graphic, width = 15, height = 10, dpi = 300)
```

